Normalization by Evaluation for Sized Dependent Types

### 2017-02-05 nbe/icfp17.tex
[ ICFP 17 ] Introduced category Data (c).

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid \tzero \mid \suc a t & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b,c & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  Similar, we use named lambda abstraction as sugar for de Bruijn.  Named abstract takes care of proper lifting of de Bruijn indices.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, \tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

  \rux{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der u\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, u : T \,b \, u
     }{u \in \{\tzero, \suc a t'\}}

@@

    }{\Gamma \der \suc a t\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,\suc a t}

@@

      \Gamma \der T = T' : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t = t' : \allT i \Size (\funT x {\Nat i} T \ann i x) \to \funT x {\Nat (i + 1)} T \ann{i+1} x
     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \ann a u}

@@

      \Gamma \der T = T' : Nat (a + 1) \to \Set[\ell] \qquad

@@

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ a \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ \ann a \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{\tzero @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc a t) @ \casett \evalsto w}
\qquad
  \rux{t\,(\lambda x.\, x@ \fixtt)\,w \evalsto w'
     }{w @ \fixtt \evalsto w'
     }{w \in \{ \tzero, \suc a u \}}
\end{gather*}

>>>>

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \tzero \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, \tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

@@

    }{\Gamma \der (\suc a t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc a t)}
  \ru{\Gamma \der c : \Nat b \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\,T %% \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der c\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, c : T \,b \, c
     }%{u \in \{\tzero, \suc a t'\}}

@@

@@

      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \qquad

@@

      \Gamma \der T = T' : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t = t' : \FixT\,T \qquad %% \allT i \Size (\funT x {\Nat i} T \,i\, x) \to \funT x {\Nat (i + 1)} T \,{i+1}\, x
     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \, a \, u}

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ a \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ \ann a \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{\tzero @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc a t) @ \casett \evalsto w}
\qquad
  \ru{t\,\ann\infty\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }%{w \in \{ \tzero, \suc a u \}}
\end{gather*}
